smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases

### 2023-07-20 manuscript/manuscript.qmd
first draft of function description chapter

For the implementation of these routine diagnostics checks, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic purpose (Table 1). In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and how the results can be interpreted. Generally, for all functions in the `smdi` package, a *dataframe* is expected as input with a format where one row stands for one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format). Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the function. Throughout all functions, the user has the option to specify a vector with the column names of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate. According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected between strata of patients with and without the partially observed covariate. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics in the dataset [@schober2019; @austin2011; @tableone]. To call the function, a user must specify the dataset th

function returns an *amsd* object which displays an aggregated summary of the average or median (depending on the analyst's choice) and the minimum/maximum range of the derived ASMDs. The *amsd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate. For a graphical visualization, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions. This function trains and fits and random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism. The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates. For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy. This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the size between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.

@@

Finally, the third group diagnostics with the `smdi_outcome()` function examines association of the missingness indicator of the partially observed covariate and the outcome under study. The function will compute both a univariate model and a model adjusted for all other covariates specified in the dataset. Currently, three outcome regression types are supported: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be displayed exponentiated or not.

>>>>

For the implementation of these routine missing data diagnostics, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic prroperties (Table 1). In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results. Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format). Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the function. Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate. According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics in the dataset [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with the corresponding minimum-maximum ASMD range. The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate. For a graphical visualization of this, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to potential MCAR or MNAR mechanisms.

The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates. The AUC values are based on the prediction made in the respective test datasets. For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy. This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the split between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.

@@

Finally, the third group diagnostics with the `smdi_outcome()` function examines association of the missingness indicator of the partially observed covariate and the outcome under study. The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset. In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated underlying missingness mechanisms [@weberpals2023]. As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the partially observed covariate. Under MAR, given that missingness can be explained by observed covariates, a potential spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed in both univariate and adjusted models.

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be displayed exponentiated or not. The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.
